Smooth and Persistent Forecasts of German GDP:\\ Balancing Accuracy and Stability\thanks{We would like Barbara Rossi and the participants at the Vienna Forecasting Workshop 2025 for their comments and suggestions.}

### 2025-08-15 Archive IWH Simon paper/15.08.2025 afetr Katja's feedback/paper.tex
up

\usepackage{graphicx} % Required for inserting images
\usepackage{tabularx}
\usepackage{natbib}
\usepackage[spaces,hyphens]{url}
\usepackage[hidelinks]{hyperref}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{hyphenat}
\usepackage{color}
%\usepackage{subfig}
% \usepackage[justification=left,hang]{caption}
\usepackage[justification=raggedright,hang]{caption}
\usepackage{subcaption}
	%\captionsetup[subfigure]{labelfont=normal,textfont=normalfont,singlelinecheck=off,justification=centering}
\usepackage[table]{xcolor} 
\usepackage{colortbl}
\usepackage[hang]{footmisc}
\usepackage{appendix}
\usepackage{float}

\geometry{left=2.25cm,right=2.25cm,top=1.5cm,bottom=2cm}
%\hyphenation{EMU} \hyphenation{CEECs}
%\setlength{\parindent}{0pt} \setlength{\parskip}{10pt} %\setlength{\abovecaptionskip}{-5pt}
%\setlength{\belowcaptionskip}{-5pt} \setlength{\floatsep}{10pt}
\clubpenalty = 10000 
\widowpenalty = 10000
\displaywidowpenalty = 10000

% \makeatletter
% \setlength{\@fptop}{5pt} %setzt den abstand der floats vom oberen Rand auf Null- 5 entspricht der gleichen Höhe wie die wo der Text anfängt
% 		%\setlength\@fptop{0\p@ \@plus 1fil}
% \setlength\@fpsep{10pt plus 2pt minus 2pt} %Abstand zwischen den Floats
% \makeatother

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\setlength{\abovedisplayskip}{2cm}
\setlength{\abovedisplayshortskip}{2cm}
\setlength{\belowdisplayskip}{0cm}
\setlength{\belowdisplayshortskip}{2cm} %Abstand zwischen Text und Formel

\title{Smooth and Persistent Forecasts of German GDP:\\ Balancing Accuracy and Stability\thanks{We would like Barbara Rossi and the participants at the Vienna Forecasting Workshop 2025 for their comments and suggestions.}}
\author{Katja Heinisch\thanks{Halle Institute for Economic Research (IWH).} \hspace{0.2cm}
Simon van Norden\thanks{HEC Montréal and CIREQ.} \hspace{0.2cm}
Marc Wildi\thanks{Zurich University of Applied Sciences (ZHAW).}}
\date{August 2025 \\
\vspace{2cm} \textit{Work in progress, not for circulation}}

@@

%Economic forecasters face a dilemma: While efficiency suggests that their forecasts should incorporate all available information, the resulting forecasts are often volatile, which may reflect greater information content or contamination by excessive ``noise''. Further, forecast users (including public and private decision makers) may be constrained by how rapidly they can adjust to changing forecasts.\footnote{For example, private or public spending decisions may be constrained by annual budgets. Similarly, central banks increasingly attempt to smooth interest rate changes to allow them to signal the future direction of policy to markets, while private firms attempt to smooth changes in dividends.}

Economic forecasters face a dilemma: while informational efficiency requires that forecasts incorporate all available data, doing so often results in forecasts that are excessively volatile.
 %Volatile forecasts may be a sign that they efficiently incorporate all available information. 
 In addition, forecasts may reflect excessive ``noise'' due to overfitting. At the same time, forecast users (including public and private decision makers) may face constraints in their speed of adjustment to changing forecasts. For example, fixed annual budgets may constrain spending, or transaction costs may limit changes in investments. Others may attempt to smooth their reactions for reasons of signalling. Important examples include dividend smoothing by corporate managers and interest rate smoothing by central banks.

Recent work by \cite{Wildi2024,Wildi2025} and \cite{McElroy2019,McElroy2020} formalipropose novel methods that optimally smooth forecasts by controlling the expected frequency of sign changes. The resulting Smooth Sign Accuracy (SSA) framework encompasses efficiently the mean-square error (MSE) as a special case and produces increasingly smooth forecasts as changes in the forecast direction (sign) are increasingly penalized.

@@

%This study proposes a methodological innovation that extends the classical direct forecasting framework by integrating dimension-reduced information extracted through Multivariate Smooth Sign Accuracy (M-SSA). The core idea is to replace raw
%, unprocessed 
%leading indicators with smoothed and filtered components derived from the M-SSA procedure, thereby improving the signal-to-noise ratio in the predictor variables used to forecast GDP. We apply this framework to well-established German economic indicators to show its ability to improve forecasts of GDP and GDP trends using a variety of sample periods and performance metrics.

This study extends this framework by replacing raw leading indicators with smoothed and filtered components, thereby improving the signal-to-noise ratio in the predictor variables. We apply this framework to well-established German economic indicators to show its ability to improve forecasts of GDP growth \citep{lehmann2021predicting,Heinisch2018bottom}  and GDP trends using a variety of sample periods and performance metrics.

@@

%Previous research often highlights the utility of direct forecasts, particularly for short-term predictions such as nowcasting and forecasts one quarter ahead. However, this paper advances the literature by showing that more sophisticated predictors—specifically, filtered components obtained via M-SSA yield superior performance, especially over medium-term horizons ranging from two to four quarters ahead.

We proceed as follows. Section \ref{sec:data} describes our data series, the sample period used, and documents their dynamic behaviour. Section \ref{sec:direct_forecast} provides benchmark forecasts using simple direct projections of GDP based on selected indicators. We demonstrate how in-sample forecast performance may improve when we ``de-noise'' the indicators prior to projection with simple smoothing filters. Section \ref{sec:mSSA} defines the M-SSA method and shows how its forecast performance compares to that of direct projections as well as how it varies with forecast horizon. 
Section \ref{sec:mSSA_component} considers the ability of our indicators to forecast German GDP growth out-of-sample using the M-SSA framework, and Section \ref{sec:revisions} investigates the temporal stability of the estimated forecasting equations. Section \ref{sec:conclusions} summarizes our results and concludes.

\section{Data and Dependence}\label{sec:data}
\subsection{Data and Publication Lags}
The analysis employs a set of well-established leading indicators to forecast quarterly German GDP \citep{Drechsel2012financial,Heinisch2018bottom}, including industrial production (IP), the ifo Business Climate Index (ifo\_c), the Economic Sentiment Indicator (ESI), and the term spread between 10-year bond yields and the 3-month EURIBOR rate (spre\_10y\_3m) covering the period 1995 to 2024.  
We take into account the publication lag of the indicators and the GDP series (Table \ref{tab:data_stucture}).\footnote{Data revisions are not considered, given that previous studies \citep{Heinisch2019} have shown that their overall effect on forecast performance is minor.} 
While GDP is available with a one-quarter delay, and IP is released with a two-month lag, business surveys and financial data are published without any lag. %All series are consequently re-aligned to correspond with the information available to forecasters, with GDP and IP variables shifted forward according to their respective lags. Table \eqref{tab:data_stucture_aligned} displays the dataset as of the end of 2024, with GDP and IP variables temporally aligned at the sample endpoint. This configuration reflects the matrix of explanatory variables accessible in Q4-2024 for the purposes of nowcasting or forecasting GDP. For illustrative purposes, an additional target column is included on the right side of the table, representing the dependent variable for nowcasting applications. In the case of an $h$-step ahead forecast, this column is shifted upwards by $h$ quarters.

\begin{table}[ht]
\caption{Ragged edge structure. As of January 2025, before publication of the GDP flash estimate.} 
\label{tab:data_stucture}
\centering
\begin{tabular}{rrrrrr}
  \hline
 & GDP & IP & ifo\_c & ESI & spr\_10y\_3m \\ 
  \hline
2024-07-01 &  & 90.700 & 85.800 & 92.500 & -1.300 \\ 
  2024-08-01 &  & 93.200 & 84.900 & 90.900 & -1.400 \\ 
  2024-09-01 &902.571  & 91.200 & 83.700 & 89.800 & -1.400 \\ 
  2024-10-01 &  & 90.800 & 84.100 & 90.700 & -1.200 \\ 
  2024-11-01 &  & 92.200 & 83.800 & 89.300 & -0.900 \\ 
  2024-12-01 &  &  & 82.800 & 86.900 & -0.800 \\ 
  2025-01-01 &  &  & 82.500 & 88.100 & -0.200 \\ 
   \hline
\end{tabular}

@@

\begin{table}[ht]
\caption{Aligned dataset. At each date, various columns show the most recent values of each predictor available to forecasters. Column 6 represents the target variable for a GDP nowcast (i.e. the values of column 1 shifted upwards by the publication lag of GDP.} 
\label{tab:data_stucture_aligned}
\centering
\begin{tabular}{rrrrrrrc}
  \hline
 & GDP & IP & ifo\_c & ESI & spr\_10y\_3m &&Target (nowcast)\\ 
  \hline
  2024-07-01 & & 93.400 & 85.800 & 92.500 & -1.300&|& \\ 
  2024-08-01 & & 90.700&  84.900 & 90.900 & -1.400&| &\\ 
  2024-09-01 & 901.623 &93.200 & 83.700 & 89.800 & -1.400&| &902.571\\ 
  2024-10-01 &  &91.200 & 84.100 & 90.700 & -1.200&| &\\ 
  2024-11-01 &  & 90.800& 83.800 & 89.300 & -0.900&| &\\ 
  2024-12-01 &902.571   & 92.200  & 82.800 & 86.900 & -0.800&| &?\\ 
    \hline
\end{tabular}

To ensure comparability and mitigate scale effects, all indicators (except the term spread, which is differenced only) are log-differenced and standardized. Additionally, extreme outliers associated with the COVID-19 pandemic are trimmed to prevent distortions in subsequent figures. 
 Monthly series are shifted (Table \ref{tab:data_stucture_aligned}) and averaged to quarterly series (Figure \ref{fig:data}), while Figure \ref{fig:data_lags} offers a close-up view of the financial crisis and the pandemic. The latter figure shows that the real-time GDP and industrial production series are synchronized at turning points, whereas the other indicators tend to lead somewhat. The challenge for forecasters is to use the latter in a multivariate approach to improve forecasting performance over that of univariate benchmarks.

\begin{figure}
  \begin{subfigure}[t]{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{./Figures/Data.pdf}
    \subcaption{Full Sample.}
    \label{fig:data}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{./Figures/data_lags.pdf}
    \subcaption{Leads \& lags during financial crisis \& Pandemic.}
    \label{fig:data_lags}
\end{subfigure}
\caption{GDP \& Indicators: standardized log-differences, trimmed to $\pm 3$ standard deviations.}
\end{figure}

@@

\subsection{Dependence}
To examine the joint dynamics of our selected variables, we estimate their cross-correlations as well as their vector moving-average (VMA) representations. The latter is a key input to the multivariate filters we develop in Sections \ref{sec:mSSA} and \ref{sec:mSSA_component}.

Figures \ref{CCF} (full sample) and \ref{CCF_wc} (omitting the Pandemic period) show the sample cross-correlation functions (CCF's) between current and lagged GDP and the two survey indicators. Both figures show strong correlations that peak when GDP is slightly lagged. Results for IP showed no such lag, while correlations for the term spread were weaker. The figures also show that the precise strengths of the correlations are influenced by the outliers associated with the COVID-19 pandemic. In the interest of model stability, we exclude data from Q4-2019 to Q4-2020 (pandemic) when modelling the dynamics for the multivariate filters developed below.

\begin{figure}[ht]
    \begin{subfigure}[htpb]{\textwidth}
        \begin{center}
            \includegraphics[height=3in, width=4.5in]{./Figures/CCF.pdf}
            \caption{Full data set from Q2-1991 to Q4-2024.\label{CCF}}
        \end{center}
    \end{subfigure}
    \\
    \begin{subfigure}[htpb]{\textwidth}
        \begin{center}
            \includegraphics[height=3in, width=4.5in]{./Figures/CCF_wc.pdf}
            \caption{Without Pandemic (omitting Q4-2019 to Q4-2020).\label{CCF_wc}}
        \end{center}
    \end{subfigure}
    \caption{Sample cross-correlation functions (CCF) showing correlations with lags of GDP growth.}
\end{figure}

We then estimated a variety of parsimonious VARMA models to capture these dynamics, selecting parsimonious VAR(1) and VAR(3) specifications.\footnote{We used the \texttt{refVARMA()} procedure in the \cite{MTSpackage} \texttt{MTS} package, which sets VARMA coefficients with t-statistics below a critical threshold to zero \citep[see][]{tsay2013multivariate}.  This led to a VAR(1) specification with a threshold of 1.5. The VAR(1) left some residual evidence of higher-order dynamics, but capturing these in our limited sample size was difficult without overfitting. We therefore explored VAR($p$) models with coefficients estimated by Elastic Net using the \cite{ElasticNet} \texttt{elasticnet} package. This led to a VAR(3) specification with 7 steps. For additional details, see \cite{ElasticNet} and the references therein.} The Figures \ref{ma_inv_BIP} and \ref{ma_inv_BIP_var3_en} show the implied responses of GDP to shocks to each indicator variable.\footnote{Shown are the MA coefficients for GDP from the vector moving-average (VMA) models obtained by inversion of the VAR(1) and VAR(3) representations respectively (i.e. the impulse-response functions (IRFs) of GDP to the correlated reduced-form shocks).}

%
%   SvN: I've added the following very brief description of Elastic-Net
%
% Linear regression models often suffer from issues such as overfitting and multicollinearity when the number of predictors exceeds the number of observations. Regularization techniques, such as ridge regression and the lasso, offer solutions by imposing penalties on model parameters. However, ridge regression maintains all predictors, while the lasso leads to sparse solutions by shrinking coefficients to zero. The elastic net estimator, introduced by \cite{ZouHastie2004JRSSB}, addresses limitations inherent in these approaches by blending their penalty functions.

@@

% where $\lambda_1$ and $\lambda_2$ are non-negative tuning parameters, and $X_i$ represents the predictor variables.
% This formulation encompasses the least-squares estimator as the special case where $\lambda_1 = 0$ and the LASSO as the special case where $\lambda_2 = 0$. Unlike the lasso, which may arbitrarily select correlated predictors, the elastic net encourages grouping of strongly correlated variables. The combined penalties stabilize estimation in cases with high-dimensional predictors.
%
%

\begin{figure}[htpb]
    \begin{subfigure}[htpb]{\textwidth}
        \begin{center}
            \includegraphics[height=3in, width=6in]{./Figures/ma_inv_BIP.pdf}
            \caption{implied by VAR(1) \label{ma_inv_BIP}}
        \end{center}
    \end{subfigure}
    \\
    \begin{subfigure}[htpb]{\textwidth}
        \begin{center}
            \includegraphics[height=3in, width=6in]{./Figures/ma_inv_BIP_var3_en.pdf}
            \caption{implied by VAR(3)\label{ma_inv_BIP_var3_en}}
        \end{center}
    \end{subfigure}
    \caption{VMA coefficients for GDP.\\Estimation with pandemic (left) and without pandemic (right).}
\end{figure}

%The models encapsulate key characteristics of the data after adjusting for publication lags. Notably, 
GDP exhibits the strongest response to shocks in the two survey indicators—ifo and ESI.\footnote{There is also weaker evidence of a response to the term spread at longer lags.} The impact of the pandemic modifies the lag structure by emphasizing shorter dependencies triggered by the alternating outliers during that specific episode. Removing the pandemic from the analysis results in smoother, somewhat longer-tailed dependencies, consistent with the observed empirical autocorrelation functions. In particular, the VAR(3) model captures slightly longer lead times and more nuanced lag patterns.\footnote{The VAR(3) model yields marginal improvements in multivariate filtering performance. However, due to its relative simplicity, this study primarily employs the VAR(1) model. The M-SSA package (accessible at \url{https://github.com/wiaidp/R-package-SSA-Predictor}) offers more advanced capabilities, including the implementation of the VAR(3) model and Bayesian VAR specifications.}

@@

\section{Forecasts with Pre-Filtering}\label{sec:direct_forecast}

In this section, we demonstrate how pre-filtering indicator variables may improve the performance of simple forecasts by removing noise. We demonstrate how in-sample forecast performance improves when we ``de-noise'' the indicators prior to projection with simple smoothing filters.\footnote{The smoothing filters are one-sided (concurrent) filters of length 31, therefore, estimation of both forecasting models omits the first 30 observations.}

For simplicity, all the models in this section are direct projections of GDP on our indicators and take the form

\begin{equation}
    y_{t+h} = \boldsymbol{\beta'} \cdot \mathbf{X}_t + \varepsilon_t
    \label{eq:direct_forecast}
\end{equation}

where
\begin{itemize}
    \item $y_{t+h}$ is the quarterly growth rate of GDP for forecast horizon $h = \{0, 1, 2, ... \}$ (labeled shifted GDP in the following )
    \item $\mathbf{X}_t \equiv [1,y_{t-1}, ip_t, ifo\_c_t, ESI_t, spr\_10y\_3m_t]'$ or their transformed counterparts and $y_{t-1}$ accounts for the publication lag of GDP.\footnote{Including lagged GDP enables the nesting of the AR forecast within the broader modeling framework.}
    \item $\hat{\boldsymbol{\beta}}$ is estimated by least-squares, omitting observations affected by the pandemic, i.e. Q4-2019 to Q4-2020.
\end{itemize}

\subsection{Unfiltered Indicators}\label{cdf}

%So-called direct forecasts are obtained by regressing the relevant indicators on forward-shifted GDP, accounting for the additional publication lag. To illustrate the concept, we rely on all indicators and full sample information, excluding observations affected by the pandemic. In-sample forecasts and the forward-shifted GDP, which serves as the target, are presented in Figures \eqref{direct_wc} (full dataset without the pandemic) and \eqref{direct_wc_financial_crisis} (financial crisis) for forecast horizons up to three quarters ahead.

@@

 \begin{subfigure}[htpb]{\textwidth}
        \begin{center}
             \includegraphics[width=0.9\textwidth]{./Figures/direct_wc_all.pdf}
            \caption{Full sample information (without pandemic)\label{direct_wc}}
        \end{center}
    \end{subfigure}

>>>>

\usepackage{graphicx} % Required for inserting images
\usepackage{tabularx}
\usepackage{natbib}
\usepackage[spaces,hyphens]{url}
\usepackage[hidelinks]{hyperref}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{hyphenat}
\usepackage{color}
\usepackage{fancyhdr}
%\usepackage{subfig}
% \usepackage[justification=left,hang]{caption}
\usepackage[justification=raggedright,hang]{caption}
\usepackage{subcaption}
	%\captionsetup[subfigure]{labelfont=normal,textfont=normalfont,singlelinecheck=off,justification=centering}
\usepackage[table]{xcolor} 
\usepackage{colortbl}
\usepackage[hang]{footmisc}
\usepackage{appendix}
\usepackage{float}

\geometry{left=2cm,right=2cm,top=1.5cm,bottom=1.5cm}

\setlength{\parindent}{0pt} %%no space at the beginning of paragraph
\setlength{\parskip}{6pt} %% space between paragraphs

\setlength{\abovecaptionskip}{0pt} %%Abstand unter der Überschrift
\setlength{\belowcaptionskip}{0pt} %%Abstand ÜBER der Überschrift
\setlength{\floatsep}{6pt}
\setlength{\textfloatsep}{6pt}
\setlength{\intextsep}{1\baselineskip}

\clubpenalty = 10000 %keine einzelnen Seiten eines Absatzes oben oder unten auf einer Seite
\widowpenalty = 10000
\displaywidowpenalty = 10000
%
\makeatletter
\setlength{\@fptop}{5pt} %setzt den abstand der floats vom oberen Rand auf Null- 5 entspricht der gleichen H?he wie die wo der Text anf?ngt
%\setlength\@fptop{0\p@ \@plus 1fil}
\setlength\@fpsep{30pt plus 2pt minus 2pt} %Abstand zwischen den Floats
\makeatother

\setlength{\abovedisplayskip}{2cm}
\setlength{\abovedisplayshortskip}{2cm}
\setlength{\belowdisplayskip}{0cm}
\setlength{\belowdisplayshortskip}{2cm} %abstand zwischen text und formel

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\title{Smooth and Persistent Forecasts of German GDP:\\ Balancing Accuracy and Stability\thanks{We would like Barbara Rossi and the participants at the Vienna Forecasting Workshop 2025 for their comments and suggestions.}}
\author{Katja Heinisch\thanks{Halle Institute for Economic Research (IWH).} \hspace{0.2cm}
Simon van Norden\thanks{HEC Montréal and CIREQ.} \hspace{0.2cm}
Marc Wildi\thanks{Zurich University of Applied Sciences (ZHAW).}}
\date{August 2025} 
%\vspace{2cm} \textit{Work in progress, not for circulation}}

@@

Economic forecasters face a dilemma: while informational efficiency requires that forecasts incorporate all available data, doing so often results in forecasts that are excessively volatile. 
In addition, forecasts may reflect excessive ``noise'' due to overfitting. At the same time, forecast users (including public and private decision makers) may face constraints in their speed of adjustment to changing forecasts. For example, fixed annual budgets may constrain spending, or transaction costs may limit changes in investments. Others may attempt to smooth their reactions for reasons of signalling. Important examples include dividend smoothing by corporate managers and interest rate smoothing by central banks.

Recent work by \cite{Wildi2024,Wildi2025} and \cite{McElroy2019,McElroy2020} formalizes this problem and propose a novel method that optimally smooths forecasts by controlling the expected frequency of sign changes. This Smooth Sign Accuracy (SSA) framework encompasses efficiently the mean-squared error (MSE) as a special case and produces increasingly smooth forecasts as changes in the forecast direction (sign) are increasingly penalized.

@@

This study extends the SSA-framework in two ways: First, replacing raw leading indicators with smoothed and filtered components, thereby improving the signal-to-noise ratio in the predictor variables. Second, we generalize SSA to a multivariate context (M-SSA), exploiting cross-sectional information among leading indicators to enhance forecast accuracy and stability. We apply this framework to well-established German economic indicators to show its ability to improve forecasts of GDP growth \citep{lehmann2021predicting,Heinisch2018bottom} and GDP trends using a variety of sample periods and performance metrics. Previous research often highlights the utility of direct forecasts, particularly for short-term predictions such as nowcasting and forecasts one quarter ahead. However, this paper advances the literature by showing that more sophisticated predictors—specifically, filtered components obtained via M-SSA yield superior performance.

@@

We proceed as follows. Section \ref{sec:data} describes our data, the sample period used, and documents their dynamic behaviour. Section \ref{sec:direct_forecast} provides benchmark forecasts using simple direct projections of GDP based on selected indicators. We demonstrate how in-sample forecast performance may improve when we ``de-noise'' the indicators prior to projection with simple smoothing filters. Section \ref{sec:mSSA} defines the M-SSA method. 
Section \ref{sec:performance_mssa} reports empirical results and shows the forecast performance compared to that of direct projections as well as how it varies with forecast horizon. 
%%%Section \ref{sec:mSSA_component} considers the ability of our indicators to forecast German GDP growth out-of-sample using the M-SSA framework.
Section \ref{sec:revisions} examines robustness and stability. Section \ref{sec:conclusions} summarizes our results and concludes. Additional Figures and Tables are presented in an Online Appendix.\footnote{\url{www.iwh-halle.de}}

\section{Data and Dependence}\label{sec:data}
\subsection{Data and Publication Lags}
Our analysis employs a set of well-established leading indicators to forecast quarterly German GDP \citep{Drechsel2012financial,Heinisch2018bottom}, including industrial production (IP), the ifo Business Climate Index (ifo\_c), the Economic Sentiment Indicator (ESI), and the term spread between 10-year bond yields and the 3-month EURIBOR rate (spre\_10y\_3m) covering the period 1991 to 2024. 
We take into account the publication lag of the indicators and the GDP series (Table \ref{tab:data_stucture}).\footnote{Data revisions are not considered, given that previous studies \citep{Heinisch2019} have shown that their overall effect on forecast performance is minor.} 
While GDP is available with a one-quarter delay, and IP is released with a two-month lag, business surveys and financial data are published contemporaneously.

\begin{table}[ht]
    \caption{Ragged edge structure.\\ Information set available to forecasters as of January 2025, before publication of the GDP flash estimate.} 
    \label{tab:data_stucture}
    \centering
    \begin{tabular}{rrrrrr}
      \hline
     & GDP & IP & ifo\_c & ESI & spr\_10y\_3m \\ 
      \hline
    2024-07-01 &  & 90.700 & 85.800 & 92.500 & -1.300 \\ 
      2024-08-01 &  & 93.200 & 84.900 & 90.900 & -1.400 \\ 
      2024-09-01 &902.571  & 91.200 & 83.700 & 89.800 & -1.400 \\ 
      2024-10-01 &  & 90.800 & 84.100 & 90.700 & -1.200 \\ 
      2024-11-01 &  & 92.200 & 83.800 & 89.300 & -0.900 \\ 
      2024-12-01 &  &  & 82.800 & 86.900 & -0.800 \\ 
      2025-01-01 &  &  & 82.500 & 88.100 & -0.200 \\ 
       \hline
    \end{tabular}
\end{table}

@@

\begin{table}[ht]
    \caption{Aligned dataset.\\
    Indicators realigned to reflect real-time availability, with the GDP column shifted by its publication lag. Column 6 represents the target variable for a GDP nowcast (i.e. the values of column 1 shifted upwards by the publication lag of GDP.} 
    \label{tab:data_stucture_aligned}
    \centering
    \begin{tabular}{rrrrrrrc}
      \hline
     & GDP & IP & ifo\_c & ESI & spr\_10y\_3m &&Target (nowcast)\\ 
      \hline
      2024-07-01 & & 93.400 & 85.800 & 92.500 & -1.300&|& \\ 
      2024-08-01 & & 90.700&  84.900 & 90.900 & -1.400&| &\\ 
      2024-09-01 & 901.623 &93.200 & 83.700 & 89.800 & -1.400&| &902.571\\ 
      2024-10-01 &  &91.200 & 84.100 & 90.700 & -1.200&| &\\ 
      2024-11-01 &  & 90.800& 83.800 & 89.300 & -0.900&| &\\ 
      2024-12-01 &902.571   & 92.200  & 82.800 & 86.900 & -0.800&| &?\\ 
        \hline
    \end{tabular}
\end{table}

To ensure comparability and mitigate scale effects, all indicators (except the term spread, which is differenced only) are log-differenced and standardized. Additionally, extreme outliers associated with the COVID-19 pandemic are trimmed to prevent distortions in subsequent figures. 
 Monthly series are aligned (Table \ref{tab:data_stucture_aligned}) and averaged to quarterly series (Figure \ref{fig:data}), while Figure \ref{fig:data_lags} offers a close-up view of the financial crisis and the pandemic. The latter figure shows that the real-time GDP and industrial production series are synchronized at turning points, whereas the other indicators tend to lead somewhat. The challenge for forecasters is to use the latter in a multivariate approach to improve forecasting performance over that of univariate benchmarks.

\begin{figure}
    \begin{subfigure}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=\linewidth]{./Figures/Data.pdf}
        \subcaption{Full Sample.}
        \label{fig:data}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=\linewidth]{./Figures/data_lags.pdf}
        \subcaption{Leads \& lags during financial crisis \& Pandemic.}
        \label{fig:data_lags}
    \end{subfigure}
    \caption{Quarterly German GDP and Indicators.\\ standardized log-differences, trimmed to $\pm 3$ standard deviations to facilitate visual inspection.}
\end{figure}

@@

\subsection{Dependence}
To examine the joint dynamics of our selected variables, we estimate their cross-correlations as well as their vector moving-average (VMA) representations. The latter is a key input to the multivariate filters we develop in Sections \ref{sec:mSSA} and \ref{sec:mSSA_component}. 
The sample cross-correlation functions (CCF's) between current and lagged GDP and the two survey indicators show strong correlations that peak when GDP is slightly lagged.\footnote{See Figures \ref{fig:CCF} (full sample) and \ref{fig:CCF_wc} (omitting the Pandemic period) in the Online Appendix.} Results for IP showed no such lag, while correlations for the term spread were weaker. However, the precise strengths of the correlations are influenced by the outliers associated with the COVID-19 pandemic. In the interest of model stability, we exclude data from Q4-2019 to Q4-2020 (pandemic) when modelling the dynamics for the multivariate filters developed below.

We then estimate a variety of parsimonious VARMA models to capture these dynamics, selecting parsimonious VAR(1) and VAR(3) specifications.\footnote{We used the \texttt{refVARMA()} procedure in the \cite{MTSpackage} \texttt{MTS} package, which sets VARMA coefficients with t-statistics below a critical threshold to zero \citep[see][]{tsay2013multivariate}. This led to a VAR(1) specification with a threshold of 1.5. The VAR(1) left some residual evidence of higher-order dynamics, but capturing these in our limited sample size was difficult without overfitting. We therefore explored VAR($p$) models with coefficients estimated by Elastic Net using the \cite{ElasticNet} \texttt{elasticnet} package. This led to a VAR(3) specification with 7 steps. For additional details, see \cite{ElasticNet} and the references therein.} The implied responses of GDP to shocks to each indicator variable are indicating that survey indicators have the strongest and most persistent impact on GDP, with pandemic data shortening the apparent lead time.\footnote{See Figures \ref{fig:ma_inv_BIP} and \ref{fig:ma_inv_BIP_var3_en} in the Online Appendix.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   SvN: I've added the following very brief description of Elastic-Net
%
% Linear regression models often suffer from issues such as overfitting and multicollinearity when the number of predictors exceeds the number of observations. Regularization techniques, such as ridge regression and the lasso, offer solutions by imposing penalties on model parameters. However, ridge regression maintains all predictors, while the lasso leads to sparse solutions by shrinking coefficients to zero. The elastic net estimator, introduced by \cite{ZouHastie2004JRSSB}, addresses limitations inherent in these approaches by blending their penalty functions.

@@

% where $\lambda_1$ and $\lambda_2$ are non-negative tuning parameters, and $X_i$ represents the predictor variables.
% This formulation encompasses the least-squares estimator as the special case where $\lambda_1 = 0$ and the LASSO as the special case where $\lambda_2 = 0$. Unlike the lasso, which may arbitrarily select correlated predictors, the elastic net encourages grouping of strongly correlated variables. The combined penalties stabilize estimation in cases with high-dimensional predictors.
%

GDP exhibits the strongest response to shocks in the two survey indicators—ifo and ESI.\footnote{There is also weaker evidence of a response to the term spread at longer lags.} The impact of the pandemic modifies the lag structure by emphasizing shorter dependencies triggered by the alternating outliers during that specific episode. Removing the pandemic from the analysis results in smoother, somewhat longer-tailed dependencies, consistent with the observed empirical autocorrelation functions. In particular, the VAR(3) model captures slightly longer lead times and more nuanced lag patterns.\footnote{The VAR(3) model yields marginal improvements in multivariate filtering performance. However, due to its relative simplicity, this study primarily employs the VAR(1) model. The M-SSA package (accessible at \url{https://github.com/wiaidp/R-package-SSA-Predictor}) offers more advanced capabilities, including the implementation of the VAR(3) model and Bayesian VAR specifications.}

@@

\section{Baseline Forecasting Approaches}
\label{sec:direct_forecast}

In this section, we evaluate the performance of simple indicator-based forecasts of German GDP growth and examine how pre-filtering can improve their accuracy by removing noise. 
Our starting point is a direct in-sample forecasting framework in which GDP growth at horizon $h$ is regressed on a set of leading indicators and lagged GDP, estimated over the pre-pandemic period.

\begin{equation}
    u_{t+h} = \boldsymbol{\beta'} \cdot \mathbf{X}_t + \varepsilon_t
    \label{eq:direct_forecast}
\end{equation}

where
\begin{itemize}
    \item $u_{t+h}$ is the quarterly growth rate of GDP for forecast horizon $h = \{0, 1, 2, ... \}$ (labeled shifted GDP in the following )
    \item $\mathbf{X}_t \equiv [1,u_{t-1}, ip_t, ifo\_c_t, ESI_t, spr\_10y\_3m_t]'$ or their transformed counterparts and $u_{t-1}$ accounts for the publication lag of GDP.\footnote{Including lagged GDP enables the nesting of the AR forecast within the broader modeling framework.}
    \item $\hat{\boldsymbol{\beta}}$ is estimated by least-squares, omitting observations affected by the pandemic, i.e. Q4-2019 to Q4-2020.
\end{itemize}

We first estimate forecasts using unfiltered indicators, which reflect the raw information available to forecasters in real time. We then apply a one-sided concurrent smoothing filter to the same indicators and demonstrate how in-sample forecast performance improves when we ``de-noise'' the indicators before projection with simple smoothing filters.\footnote{The smoothing filters are one-sided (concurrent) filters of length 31, therefore, estimation of both forecasting models omits the first 30 observations.} 
Comparing these two designs allows us to isolate the effect of pre-filtering in a straightforward, transparent way.

\subsection{Direct Forecasts with Unfiltered Indicators}\label{sec:cdf}

As a benchmark, we estimate direct projections of GDP growth at horizons $h=0$ to $h=6$ quarters ahead using all indicators without filtering.   
Figures \ref{fig:direct_wc} and \ref{fig:direct_wc_financial_crisis} compare actual and fitted values for GDP at forecast horizons $h = 0, \ldots, 3$, using the four indicators plus lagged GDP. 
%At short horizons, forecasts track turning points relatively well but exhibit considerable volatility. With increasing horizon, the forecasts become smoother; however, this is achieved at the expense of timeliness---turning points, i.e., peaks and troughs, are increasingly delayed relative to the target series GDP. 
At short horizons, forecasts track turning points relatively well. As the forecast horizon increases, the forecasts become increasingly damped (zero-shrinkage) and delayed, with peaks and troughs progressively shifted to the right relative to the target series GDP.

\bigskip
\begin{table}[h]
\caption{Statistical Significance of Direct Forecasts.\\
p-values for $H_0: \boldsymbol{\beta} = 0$ in eq.~\eqref{eq:direct_forecast} based on HAC-adjusted Wald test.\\Estimation over full sample (without pandemic.)} 
\label{tab:pvaluedhp}
\centering
\begin{tabular}{lrrrrrr}
  \hline
  & h= 1 & h= 2 & h= 3 & h= 4 & h= 5 & h= 6 \\ 
  \hline
  Unfiltered & 0.000 & 0.001 & 0.910 & 0.509 & 0.091 & 0.178 \\ 
HP-C filtered & 0.001 & 0.036 & 0.046 & 0.049 & 0.012 & 0.022 \\ 
   \hline
\end{tabular}

Table \ref{tab:pvaluedhp} reports p-values for tests of the joint null hypothesis that all coefficients on the indicators are zero (HAC-adjusted Wald test). 
Additionally, these results confirm that the predictive power of the unfiltered indicators declines sharply beyond $h=2$, with significance levels falling to insignificance for horizons beyond two quarters.%\footnote{These findings are confirmed by the relative root mean-square errors comparing direct forecasts against the mean benchmark, see Table \ref{tab:rRMSE_mSSA_direct_mean_without_covid8} in the Appendix.}
 This finding motivates the use of pre-filtering to extract a clearer signal from the indicator set.

@@

 \begin{subfigure}[htpb]{\textwidth}
        \begin{center}
             \includegraphics[width=0.9\textwidth]{./Figures/direct_wc_all.pdf}
            \caption{Full sample information (without pandemic)\label{fig:direct_wc}}
        \end{center}
    \end{subfigure}
